---
title: "Air raids on the UK: understanding the effect of time on casualties"
subtitle: "How nighttime can influence the destructiveness of warfare"
author: 
  - Dennis Netchitailo
thanks: "Code and data are available at: 
[https://github.com/dennisnetchitailo/STA304P3](https://github.com/dennisnetchitailo/STA304P3)."
date: 02 December 2024
date-format: long
abstract: "With wars raging across the world, notably, between Russia and Ukraine, and aerial bombardment using aircraft or missiles a major cause of civilian casualties far from the frontlines, understanding, based on historical data, whether either nighttime or daytime attacks are more damaging would allow policymakers to better allocate resources when dealing with the treat of aerial attack. This study models the German bombing campaign of Britain during the Second World War by analyzing the data collected by the Home Security War Room from 1939 to 1945 that summarised air raid damage. I use a multiple linear regression model to estimate the impact of the time an attack occurred on the resulting casualties."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
library(here)
library(readr)
library(usethis)
library(dplyr)
library(testthat)
library(rstanarm)
library(ggplot2)
library(brms)

```


# Introduction

////Overview paragraph////
On September 1st, 1939, the German armed forces crossed the Polish border. The United Kingdom declared war on Germany on September 3rd, 1939. 
The Ministry of Home Security was the British government department established in 1939 to direct the response to German air raids. Part of its responsibility was tracking data related to air raids. This data  


////Estimand paragraph////
The estimand is the number of casualties sustained during an attack. I use the Home Security data with a multiple linear regression model to estimate the difference in casualties during nighttime air raids compared to daytime air raids 

////Results paragraph////

////Why it matters paragraph////

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....






# Data {#sec-data}

## Overview

This study uses the [@airraids]

I used the statistical programming language R [@citeR] for data cleaning and analysis. The R packages used were **`tidyverse`** for data manipulation and visualization [@tidyverse], **`here`** for file path management [@here], **`lubridate`** for date handling [@lubridate], **`arrow`** for data storage and processing [@arrow], **`testthat`** for unit testing [@testthat], **`readr`** for reading structured data [@readr], **`dplyr`** for data manipulation [@dplyr], **`stringr`** for string operations [@stringr], **`ggplot2`** for creating visualizations [@ggplot2], **`caret`** for modeling [@caret], **`kableExtra`** for creating tables [@kableExtra], and **`modelsummary`** for summarizing models and results [@modelsummary].

.... Our data [@airraids].... Following @tellingstories, we consider...

arrow /
here / 
readr /
usethis
dplyr / 
testthat / 
rstanarm
ggplot2 /
brms

## Measurement
	
The measurement process refers to the attempt to explain real world events through data, such as an individual act of warfare, as represented by the use of a self-propelled (missile) or gravity released explosive (bomb) to strike land or sea target, into numerical entries describing this event. Each entry in the dataset represents a bombing incident that took place in the UK during WW2. 
	
**Country**: The country where an incident occurred. This includes England, Scotland, Wales, Northern Ireland, and the Channel Islands. 
**Location**: 
**Civil Defense Region**: A region defined by the military for purposes of organization. The United Kingdom was divided into 13 regions. 
**Coordinates: Longitude and Latitude**: The location of an incident as recorded by the mercator projection.  
**Date: Start & End**: The date an incident started and ended at a specific location. 
**Time** The time of day an incident occurred. Day is defined as being between 6AM and 6PM. Night is defined as being between 6PM and 6AM. 
**Casualties: Killed, Injured, Total**: The number of people killed or injured due to an incident. Total casualties is the sum of the number killed and injured in the same incident. 

## Data Cleaning
The two datasets that were used to make the single dataset that provided the data for the analysis were cleaned to ensure accuracy and consistency.  Missing values were dealt with by removing them from the analysis. Variables were converted into logical data types. Unneeded columns were removed. The two cleaned datasets were unified into one. 



## Outcome variables

The outcome variable is **`total_casualties`**. This is the primary dependent variable that the model is designed to predict. It represents the number of people killed or injured by air raids. The model seeks to identify the variables affecting this number. 

@fig-monthdistribution visualizes the distribution of casualties from 1939 to 1945 on a monthly basis in a histogram.

```{r}
#| label: fig-monthdistribution
#| fig-cap: "Distribution of casualties in the UK caused by air raids throughout the war."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create the histogram plot
ggplot(combined_data, aes(x = factor(start_month), y = total_casualties, fill = as.factor(start_year))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(name = "Month") +
  scale_y_continuous(name = "Total Casualties") +
  labs(title = "Total Casualties by Month (1939-1945)", fill = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
A large number of casualties occurred in the last months of 1940 and the first months of 1941. This coincides with "The Blitz", the name given to the most intense period of air raids on the UK, following the defeat of France in June 1940. The number of casualties rises month to month following the defeat of the Allies in western Europe, as Germany could concentrate its airforce on bombing the United Kingdom.

```{r}
#| label: fig-yeardistribution
#| fig-cap: "Distribution of casualties in the UK caused by air raids throughout the war."
#| echo: false
#| eval: true
#| warning: false
#| message: false
# Create the histogram plot
ggplot(combined_data, aes(x = factor(start_year), y = total_casualties)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  scale_x_discrete(name = "Year") +
  scale_y_continuous(name = "Total Casualties") +
  labs(title = "Total Casualties by Year (1939-1945)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```
@fig-yeardistribution shows that 

summary_data

```{r}
#| label: fig-yeardistribution
#| fig-cap: "Distribution of casualties in the UK caused by air raids throughout the war."
#| echo: false
#| eval: true
#| warning: false
#| message: false
# Create the histogram plot
# Create the dual-axis plot
ggplot(combined_data, aes(x = factor(start_year))) +
  geom_bar(aes(y = total_casualties), stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = total_incidents * (max(total_casualties) / max(total_incidents)), group = 1), 
            color = "darkred", size = 1) +
  geom_point(aes(y = total_incidents * (max(total_casualties) / max(total_incidents))), 
             color = "darkred", size = 2) +
  scale_y_continuous(
    name = "Total Casualties",
    sec.axis = sec_axis(~ . * (max(combined_data$total_incidents) / max(combined_data$total_casualties)), 
                        name = "Number of Incidents")
  ) +
  labs(title = "Total Casualties and Number of Incidents by Year (1939-1945)", 
       x = "Year") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.title.y.right = element_text(color = "darkred"),
    axis.text.y.right = element_text(color = "darkred")
  )
```



Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.


The **predictor variables** are the factors that are assumed to influence the number of casualties. 

1. **Time ('time_binary')**. This variable represents whether an incident occurred during daytime, defined as between 6AM and 6PM, or nighttime, defined as between 6PM and 6AM. The assumption here is that the response time of medical, military, and police services differs at night, given more people at rest (implying worse response time), but also less traffic (implying faster response time). This is the key variable being assessed, given 

2. **Country ('country')** is a factor, because each country's cultural, social, and economic conditions can influence how much damage an air raid has by influencing factors such as the the abundance of underground shelter (such as the London Underground) or the quality of building construction (making a structure more or less susceptible to collapse). 

3. **Date ('start_year') and ('start_month')** is a factor because over time, technology and capabilities changed for both sides in the war. Equally, seasonal effects may have played a role. For instance, in wintertime, it is possible that there could have been cases of frostbite caused by the destruction of shelter and heating infrastructure. 

4. **Civil Defense Region ('civil_defense_region')**: is a factor because different military regions differed in their capabilities to prevent and to respond to air raids, as well as in intensity of targeting by the German airforce.  


# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("STA304P3/models/refined_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


